\chapter*{Preface}
\thispagestyle{empty}

This is my Master thesis in Computer Science,
specialisation \emph{Formal Methods and Software Verification},
submitted at the \emph{VU University Amsterdam} (or just \emph{Vrije
  Universiteit} if you like).

It describes the formalisation of some of the \emph{infinitary
  rewriting} theory in the \Coq proof assistant.
% (don't worry, you'll learn what they are if you make it to
%page~\pageref{LastPage}).
Together, the formalisation and thesis make
for a 36 ECTS project, roughly equivalent to six months of work,
concluding my Master study.

So I would like to say to myself, \emph{welcome to the real world!}
But not before some much deserved \emph{thank-you}'s (and the
remainder of this thesis).


\section*{Acknowledgements}

First of all, I would like to thank my supervisor, Dimitri
Hendriks. Almost a year ago, he coined the title of this thesis, and
has since spent many hours fixing my \Coq code, or otherwise
`supervising' me. The formalisation of infinite terms described in
Section~\ref{sec:terms}, for example, is in large part due to
Dimitri.

I also thank Roel de Vrijer, as second reader of my thesis, but also
as the one who introduced me to the worlds of formal methods and
logic---two central themes in my Master study.

Though not directly in this final project, Femke van Raamsdonk was
involved in many of my curricular activities, be it as lecturer,
supervisor, or employer, but always in a friendly and colleague-like
way. Thanks!

The work described in this thesis was also at least partially enabled
by the following people. So thank you. Peter Hancock for his notes on
tree ordinals and further personal communication. Vincent van Oostrom
for a fruitful morning session and useful comments on the embedding of
rewrite sequences. Clemens Grabmayer for elaborating on the infinitary
unique normal form property. Adam Chlipala for his recursive vector type and
other (sometimes extremely fast) answers on the \Coq mailing
list. Bruno Barras for a hint on proving a lemma on ordinals, through
that same mailing list. Matthieu Sozeau for assisting me on problems
with a \Coq tactic. And of course J\"org Endrullis for joining Dimitri
and me on our coffee/chocolate/tea breaks, but also for many insightful
comments on my project.

%Jeffrey Harris for pointing out Hancock's notes.

\thispagestyle{empty}

\begin{flushright}
  Martijn Vermaat\\
  Amsterdam\\
  August 2010
\end{flushright}
