\documentclass[a4paper,11pt]{article}
\usepackage[english]{babel}
\usepackage{a4}
%\usepackage{a4,fullpage}
%\usepackage{amsmath,amsfonts,amssymb}
%\usepackage{fancyhdr}
\usepackage{listings}


% Use sans serif font for body text
%\renewcommand{\familydefault}{\sfdefault}

\pagestyle{headings}


\title{Concrete Syntax for Meta-Programming}

\author{Martijn Vermaat\footnote{{\bf Important notice:}
    The research presented in this paper
    was \emph{not} done by Martijn Vermaat, but by the Software
    Technology Group at Utrecht University, most notably Eelco
    Visser.
    The purpose of this paper was only to server as an exercise
    in academic writing. This was part of the course Scientific
    Communication in Computer Science at the Vrije Universiteit.
    Original papers presenting the research described in this
    paper include \cite{Vis02,Brav04}.
} \footnote{E-mail: \texttt{mvermaat@cs.vu.nl}, homepage: \texttt{http://www.cs.vu.nl/\~{}mvermaat/}}}
\date{\today}

\makeindex


\begin{document}

\maketitle


\lstset{
  numbers=none,
  basicstyle=\footnotesize\ttfamily,
  frame=tb,
  language=Pascal,
  captionpos=b,
  xleftmargin=1em,
  xrightmargin=1em,
  aboveskip=1em,
  belowskip=1em
}


\begin{abstract}
In meta-programming, object programs are analysed, transformed, and
generated. In order to allow guaranties about the syntactical correctness to be
made and powerfull techniques to be employed in the manipulation of
object code, meta-programming systems use abstract syntax representations
of object languages.

However, the notations provided for manipulating abstract syntax are
too verbose and difficult to read for many applications. We present
a method for the embedding of concrete syntax for object languages in
the meta-language to eliminate the conceptual distance between the
notations for manipulating abstract syntax and the traditional concrete
syntax of object languages. This method is generic and can thus be
applied to arbitrary meta-languages and object languages.
\end{abstract}


\section{Introduction}\label{sec:introduction}

Meta-programming is concerned with analysis, generation, and transformation
of object programs. In this setting, the meta-language provides constructs
to manipulate code fragments of the object language. Meta-languages employ
standard data structures for the representation of object programs. Typically,
these data structures are used to represent abstract syntax trees rather
than textual (concrete) syntax.

The use of abstract syntax enables meta-programming environments to make
guaranties about well-formedness and type-correctness of generated programs.
Furthermore, abstract syntax representations allow for high-level techniques
to be employed in analysis and composition of code fragments.

However, there are many domains of meta-programming where the conceptual gap
between the concrete syntax representation of object programs and the constructs
to manipulate the abstract syntax representation of these programs is greater
than one should desire.

The techniques we present allow meta-programming systems to use abstract
syntax representations of object code, while giving the meta-programmer the
possibility of specifying the same object code using the concrete syntax of
the object language. This entails a more natural way of meta-programming with
higher readability of meta-programs, while still having all the bennefits of
abstract syntax representations of object programs.

\paragraph{}

We proceed as follows. In section~\ref{sec:motivation} we introduce and explain
the problem domain. In section~\ref{sec:related} we give a brief overview of
previous and related work. The actual implementation of our solution is
discussed in section~\ref{sec:realization}. We conclude in section~\ref{sec:discussion}.


\section{Motivation}\label{sec:motivation}

Systems for meta-programming are in widespread use. They are being used for
the analysis of existing application code (e.g. to find bugs, to prove
correctness). Meta-programming is also concerned with transforming existing
code (e.g. optimization, translation). Furthermore, automatically generating
program code is an important application of meta-programming (e.g. compilers,
relational database interfacing).

\paragraph{}

Systems for meta-programming use data structures for the representation of
object code fragments. One possible approach is to use the string type of the
meta-language for storing fragments of the object program in concrete syntax.
Composition of fragments is simply done by string concatenation, whereas
analysis and deconstruction of fragments is much harder.

A short characterization of the approach is the following. Specifying object
fragments is done in a natural way, using concrete syntax. Reading
meta-programs is easy because of this concrete syntax. Manipulating object
code can be much harder, because no structure is preserved. No guarantees
whatsoever can be made about the correctness of the generated code (i.e.
syntactical correctness, type-correctness).

\paragraph{}

Another approach is to use a datatype to store an abstract syntax representation
of the object program. The use of abstract syntax allows high-level data
manipulation techniques to be employed in the meta-language in order to
manipulate the object program. For example, object-oriented languages such
as Java provide methods to store abstract syntax trees as object hierarchies
and techniques to compose and decompose these hierarchies, while functional
programming languages support algebraic data types in combination with pattern
matching (e.g. Haskell, and the ML family of languages).

Compared to using strings of object code fragments, abstract syntax
representations allow the meta-programming system to make guaranties about the
well-formedness of generated object programs in addition to providing much more
powerfull techniques for further manipulation of these programs.

A negative aspect of this approach is that the meta-programmer has no natural
way of specifying object code. This has a negative effect on both the
construction and the readability of meta-programs. As an example, consider the
following arithmetic expression in an imaginary object language.
\begin{lstlisting}[title=Example expression in concrete syntax]
(1 + 7) * i - 1
\end{lstlisting}
A typical abstract syntax representation of this fragment in a meta-language
with algebraic data types can look like the following.
\begin{lstlisting}[title=Example expression in abstract syntax]
Sub(Mul(Add(IConst(1), IConst(7)), Id('i')), IConst(1))
\end{lstlisting}
It is clear that in this example the concrete syntax representation of the
object code is much easier to read than the abstract syntax representation in
the meta-language. Although it should be noted that in some cases the abstract
syntax representation can be more concise, in general it tends to become more
painful to manage abstract syntax as the size of object code fragments grows.

\paragraph{}

In this paper we present a method for the embedding of concrete syntax of
object languages in the syntax of meta-languages (thus thereby extending the
meta-language) and the assimilation of this embedded concrete syntax into the
meta-language. The embedding of an object language in a meta-language is done
by combining their syntax definitions. Assimilation of the embedded concrete
object code fragments is done by program transformation. Concrete object
syntax is transformed to the standard meta-language constructs for
constructing and manipulating abstract object syntax resulting in a pure
meta-program that can be compiled in the normal way.


\section{Related Work}\label{sec:related}

\subsection{User-definable Syntax}

User-definable syntax provides a way for the programmer to define certain
syntactical constructs in the language. A simple example of this are
user-definable infix operators as present in languages like Prolog and
Haskell. However, although sometimes very useful, this extending of syntax
is limited in its applications. More powerfull is the definition of syntax
in the algebraic specification formalism ASF+SDF \cite{Berg89, Deur96} in which the programmer
defines all syntactical constructs itself.

Experiments have been done with dynamically extending the syntax of a
language at parse-time \cite{Weis93, Card94, Brab02} by including declarations of syntax extensions in
the program itself. It is our opinion that this complicates the parsing
process too much while the bennefits of dynamically extending syntax within
a program file are not clear in the application of meta-programming.


\subsection{Syntax Macros}

Syntax macros \cite{Leav66} are a technique to define syntactic abstractions over code
fragments. A typical syntax macro consists of a syntax definition that
accepts structured code fragments as arguments on invocation. The result
after invocation is a new code fragment constructed by the syntax macro.

Although syntax macros can provide useful abstractions, the syntax for
invocations is always fixed and usually starts with a macro delimiter or
identifier. Also, syntax macros do not abstract over the lexical syntax
of the language which our method does.

The rewriting (assimilation) of syntax macro invocations to the host
language in different syntax macro systems differs in expressivity, but
even the most expressive systems \cite{Weis93, Sha96} require non-terminals in the user-defined
syntax to map on a fixed non-terminal in the host language. The method
we present does not have this restriction as it completely seperates syntax
definition and assimilation.


\section{Realizing Concrete Syntax}\label{sec:realization}

In this section we will present our method for the embedding of concrete
syntax and assimilation to the meta-language. To illustrate the techniques,
the embedding of the Tiger language in the Stratego language will be used
as a running example.


\subsection{Syntax Definition}

Traditional systems for meta-programming need at least a syntax definition
of the object language combined with a parser and a pretty printer. In this
setting, object programs can be parsed to an abstract syntax representation
in the meta-language and this representation can be pretty-printed to
concrete object language code, usually after modifications by the meta-program.

Our proposed method consists of the following. The syntax definitions of the
meta-language and the object language are combined to form a syntax definition
of the meta-language extended with the embedding of the object language.
Meta-programs can be written in this syntax (and thus using concrete syntax
for object code) and are transformed to programs in the traditional syntax of
the meta-language. The transformation rewrites fragments of concrete object
code to constructs in the meta-language working on an abstract syntax representation
of the object code.

\subsubsection*{SDF and SGLR}

For syntax definition, we use the Syntax Definition Formalism SDF \cite{Brand02, Visser97b} which combines
lexical and context-free syntax in a declarative and modular system. SDF definitions
are declarative in that it that they are not operational implementations of a
specific parser or parsing algorithm. Furthermore, SDF provides constructs to
combine syntax definitions and this is essential for our application. For this SDF
provides automatic renaming and importing of external syntax definitions, thereby
eliminating the need to write combined definitions entirely by hand. Because
there is no proper subclass of context-free grammars that is closed under union
SDF allows the full class of context-free grammars to be defined. Describing
both lexical and context-free syntax in a single definition is important as it
allows combining languages with completely different lexical syntax.

We employ the scannerless generalized LR parsing algorithm as implemented in
SGLR \cite{Visser97}. The algorithm is scannerless because it does not include a separate lexical
scanner (which is important for reasons stated above). Generalized LR parsing is
an adaptation of LR parsing that results in a parse forrest rather than a single
parse tree. Multiple LR parsing states will be maintaned in parallel when multiple
productions can be choosen. Generalized LR and Early parsing are the only existing
algorithms that allow the full class of context-free grammars as is dictated by
the use of SDF.

\subsubsection*{Tiger in Stratego}

The following code fragment is a condensed version of the syntax definition of
Tiger embedded in Stratego. The definition uses existing syntax definitions
for Tiger (\texttt{Tiger}) and Stratego (\texttt{Stratego}). Non-terminals
present in both languages are renamed in the imported Stratego syntax
definition and selected non-terminals of the Tiger syntax together with delimiter
syntax are used to produce ATerm sorts. If the lexical syntax of the two
languages is sufficiently different, explicit delimiters can be eliminated
entirely.

\begin{lstlisting}[title=Combination of syntax definitions of Stratego and Tiger]
module StrategoTiger
imports Tiger
imports Stratego [ID      => StrategoID
                  Var     => StrategoVar
                  StrChar => StrategoStrChar]
exports
  context-free syntax
    '[[' Dec    ']]'  -> Term  {cons('ToTerm')}
    '[[' FunDec ']]'  -> Term  {cons('ToTerm')}
    '[[' Exp    ']]'  -> Term  {cons('ToTerm')}
\end{lstlisting}


\subsection{Assimilation}

We use the Stratego language for the assimilation of embedded object code fragments
in concrete syntax into the meta-language by transformation. Stratego is a
transformation language that itself supports meta-programming with concrete object
syntax implemented using the method described here. Abstract syntax of object code
in Stratego is stored in the ATerm format \cite{Brand00}, a generic format for describing
structured data comparable to but more concise than the XML format.

Assimilation of embedded fragments can be done by a generic transformation to the
abstract syntax representation of the meta-language. This transormation can be
automatically generated from the combined syntax definition of the meta-language
and the object language. Indeed, the Stratego language uses this technique to
assimilate arbitrary object languages into the core Stratego language by rewriting
embedded concrete code fragments to the corresponding ATerm constructs.

Another approach is defining a custom transformation for assimilation, rather than
generating one automatically from the syntax definition. This allows one to exploit
the domain-specific characteristics of the embedding and thus to rewrite the
concrete syntax to more direct or efficient representations in the meta-language.

\subsubsection*{Assimilation of Tiger in Stratego}

Using the embedding of Tiger in Stratego as defined in a simplified form in the
previous subsection, we can write the following Stratego rule.
\begin{lstlisting}[title=Program fragment of Tiger in Stratego]
[[ x:= let ds in ~* es end ]] -> [[ let ds in x := (~* es) end ]]
\end{lstlisting}
By applying a generic transformation on the concrete syntax fragments to Stratego
code with ATerm constructs, this results in the folloing abstract syntax tree.
\begin{lstlisting}[title=Translated program fragment of tiger in Stratego]
Rule(ToTerm(Assign(Var(meta-var('x')),
                   Let(meta-var('ds'),FromTerm(Var('es'))))),
     ToTerm(Let(meta-var('ds'),
                [Assign(Var(meta-var('x')),
                        Seq(FromTerm(Var('es'))))])))
\end{lstlisting}


\subsection{Meta-Variables}

Concrete object code fragments embedded in the meta-program may need to refer to
meta-language variables (typically having as value another object code fragment).
This is called escaping and can be done by using a fixed syntactical construct
for meta-language variables inside object code.
However, it would be preferable if we could just write the name of a meta-variable
in concrete syntax fragments. This can be achieved by using SDF variable
declarations where we define (patterns for) identifiers that should be interpreted
as meta-variables by the parser.

Using SDF variable declarations, the identifiers \texttt{f}, \texttt{xs}, and \texttt{e}
in the following term in Stratego extended with concrete syntax for Tiger are not
Tiger variables, but meta-variables (i.e. denoting Tiger terms at the Stratego
level).
\begin{lstlisting}[title=Implicit meta-variables]
[[ function f(xs) = e ]]
\end{lstlisting}


\section{Discussion}\label{sec:discussion}

\subsection{Evaluation}

The method we presented allows for meta-programming with concrete object syntax.
This means a more natural way to write meta-programs while still having all the
bennefits of traditional meta-programming with abstract object syntax.

Compared to syntax macros, our method is more flexible in that it allows for the
embedding of arbitrary syntax, including lexical syntax, whereas syntax macros
typically only provide limited user-definable syntax and are always restricted
to the lexical syntax of the host language.

We think user-definable syntax in the sence of ASF+SDF is an interesting
technique. Indeed, Stratego grew out of experience with ASF+SDF. Systems for
dynamically extensible syntax complicate the parsing process considerably and
provide more freedom in extending syntax within program files than is needed
for meta-programming in our view. Our method separates embedded syntax
definitions from the actual program code, providing syntax definition on a
per-file basis.


\subsection{Conclusions and Future Work}

In this paper we presented a generic method for embedding an arbitrary object
language in an arbitrary meta-language. Needed for this are syntax definitions
for both languages, a combined syntax definition and a rewriting of object code
fragments to meta-language constructs. The techniques discussed are all 
available today and proven in practice by active use.

The embedding and assimilation of object languages are clearly separated from
the actual meta-programming and thus allow for a strict separation of roles.
The language engineer can define the embedding and assimilation, while the
meta-programmer does not need to have any knowledge of this process. For
meta-language compilers that accept abstract syntax as input, the pretty-printing
phase after assimilation can be discarded. For an even more integrated environment,
assimilation of the object language could be integrated in the meta-language
compiler. 

\paragraph{}

For future work, the methods discussed can be applied to syntactically extending
host languages in general rather than the use of concrete syntax in
meta-programming. Interesting applications include providing domain specific
syntax for library interfaces, embedding of concrete syntax for regular epressions
in mainstream programming languages to allow syntax checking at parse-time, and
generation of XML using the concrete syntax of XML, among others.


\begin{thebibliography}{99}

\bibitem{Berg89}J. A. Bergstra, J. Heering, and P. Klint, editors. \emph{Algebraic
Specification}. ACM Press Frontier Series. The ACM Press in co-operation with Addison-Wesley,
1989.

\bibitem{Brab02}C. Brabrand and M. I. Schwartzbach. Growing languages with metamorphix
syntax macros. In \emph{PEPM'02}, 2002.

\bibitem{Brand00}M. G. J. van den Brand, H. de Jong, P. Klint, and P. Olivier. Efficient
annotated terms. \emph{Software, Practice \& Experience}, 30(3):259-291, 2000.

\bibitem{Brand02}M. G. J. van den Brand, J. Scheerder, J. Vinju, and E. Visser.
Disambiguation filters for scannerless generalized LR parsers. In N. Horspool, editor,
\emph{Compiler Construction (CC'02)}, volume 2304 of \emph{Lecture Notes in Computer
Science}, pages 143-158, Grenoble, France, April 2002. Springer-Verlag.

\bibitem{Brav04}M. Bravenboer, E. Visser. Concrete Syntax for Objects.
Domain-Specific Language Embedding and Assimilation without Restrictions. In Douglas
C. Schmidt (ed.) \emph{Proceedings of the 19th ACM SIGPLAN conference on
Object-Oriented Programming, Systems, Languages, and Applications (OOPSLA'04).}
Vancouver, Canada. October 2004.

\bibitem{Card94}L. Cardelli, F. Matthes, and M. Abadi. Extensible syntax with lexical
scoping. SRC Research Report 121, Digital Systems Research Center, Palo Alto,
California, Februari 1994.

\bibitem{Deur96}A. van Deursen, J. Heering, and P. Klint, editors. \emph{Language Prototyping.
An Algebraic Specification Approach}, volume 5 of \emph{AMAST Series in Computing} World
Scientific, Singapore, September 1996.

\bibitem{Leav66}B. M. Leavenworth. Syntax macros and extended translation. \emph{Communications
of the ACM}, 9(11):790-793, November 1966.

\bibitem{Sha96}A. Shalit. \emph{The Dylan reference manual: the definitive guide
to the new object-oriented dynamic language}. Addison Wesley Longman Publishing
Co., Inc., 1996.

\bibitem{Visser97}E. Visser. Scannerless generalized-LR parsing.
Technical Report P9707, Programming Research Group, University of Amsterdam, July 1997.

\bibitem{Visser97b}E. Visser. \emph{Syntax Definition for Language Prototyping.} PhD
thesis, University of Amsterdam, September 1997.

\bibitem{Vis02}E. Visser. Meta-programming with concrete object syntax. In D. Batory,
C. Consel, and W. Taha, editors, \emph{Generative Programming and Component Engineering
(GPCE'02)}, volume 2487 of \emph{Lecture Notes in Computer Science}, pages 299-315, Pittsburgh,
PA, USA, October 2002. Springer-Verlag.

\bibitem{Weis93}D. Weise and R. F. Crew. Programmable syntax macros. In \emph{Proceedings
of the ACM SIGPLAN '93 Conference on Programming Language Design and Implementation
(PLDI'93)}, Albuquerque, New Mexico, June 1993.

\end{thebibliography}


\end{document}





% Meta-Programming with Concrete Object Syntax
%
% 1 Introduction
% 2 Abstract Syntax vs Concrete Syntax
%   2.1 Syntax Definition
%   2.2 Example: Instrumenting Programs
%   2.3 Concrete vs Abstract
% 3 Implementation
%   3.1 Extending the Meta-Language
%   3.2 Meta-Variables
%   3.3 Meta-Explode
% 4 Generalization
% 5 Discussion
%   5.1 Syntax Definition and Parsing
%   5.2 Desugaring Patterns
%   5.3 User-definable Syntax
%   5.4 Syntax Macros
% 6 Conclusions

% Concrete Syntax for Objects
%
% 1 Introduction
% 2 Concrete Syntax for Objects
%   2.1 Code Generation
%   2.2 XML Document Generation
%   2.3 Graphical User-Interface Construction
%   2.4 Other Applications
% 3 Realizing Concrete Syntax
%   3.1 Embedding and Assimilation
%   3.2 Java with Swul
%   3.3 Java with XML
%   3.4 Java with Java
% 4 Syntax Definition
%   4.1 SDF Overview
%   4.2 The Importance of Modularity
%   4.3 The Importance of Scannerless Parsing
% 5 Discussion
%   5.1 Previous Work
%   5.2 Related Work
%   5.3 Future Work
% 6 Conclusions
