\documentclass{beamer}

\usepackage[english]{babel}
\usepackage{listings}
\usepackage{beamerthemesplit}
\setbeamertemplate{background canvas}[vertical shading][bottom=red!10,top=blue!10]
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{headline}{}
\usetheme{Warsaw}
\useinnertheme{rectangles}

\colorlet{darkred}{red!80!black}
\colorlet{darkblue}{blue!80!black}
\colorlet{darkgreen}{green!80!black}

% By Arthur van Dam
\lstdefinelanguage{SDF}%
{otherkeywords={->,>,\{,\},;},%
  morekeywords=[1]{bla,sorts,definition,module,context,free,syntax,lexical,prefer,reject,avoid,%
    priorities,exports,imports,cons,right,left,assoc,bracket,variables,restrictions,bla}%
    sensitive,%
    morecomment=[l]{\%\%},%
    morestring=[b]"%
}[keywords,comments,strings]%

\lstset{
  numbers=none,
  basicstyle=\footnotesize\ttfamily,
  frame=tb,
  language=Java,
  captionpos=b,
  xleftmargin=1em,
  xrightmargin=1em,
  aboveskip=1em,
  belowskip=1em
}


\title{Concrete Syntax for Meta-Programming}

\author{Martijn Vermaat}
\institute{mvermaat@cs.vu.nl\\
http://www.cs.vu.nl/\~{}mvermaat/}
\date{Scientific Communication in Computer Science\\
May 25, 2005}


\begin{document}


\frame{\titlepage}


\frame{

  \frametitle{Concrete Syntax for Meta-Programming}

  \tableofcontents

}


\section{Introduction}

\frame{

  \frametitle{Meta-Programming}

  Meta-program:
  \begin{quote}
    ``Program that manipulates other programs''
  \end{quote}

  \uncover<2->{

    \begin{block}{Meta-programming is about}
      \begin{itemize}
        \item Source code analysis.
        \item Code generation.
        \item Program transformation.
      \end{itemize}
    \end{block}

  }

}


\begin{frame}

  \frametitle{Concrete syntax for manipulated language}

  \begin{block}{Manipulating code fragments}
    \begin{itemize}
      \item String literals are bad.
      \item Abstract syntax trees are good.
      \item ASTs are too verbose.
    \end{itemize}
    Conclusion: {\bf concrete syntax}
  \end{block}

\end{frame}


\section{Motivation}

\frame{

  \frametitle{Concrete Syntax for Meta-Programming}

  \tableofcontents[currentsection]

}


\begin{frame}

  \frametitle{Meta-Language and Object Language}

  \begin{block}{Meta-language}
    \begin{itemize}
      \item Language of the meta-program.
      \item Example: {\bf Javascript}
    \end{itemize}
  \end{block}

  \begin{block}{Object language}
    \begin{itemize}
      \item Language of the manipulated code.
      \item Example: {\bf XML}
    \end{itemize}
  \end{block}

\end{frame}


\begin{frame}[fragile]

  \frametitle{Example: Generating XML in Javascript}

  Generate this XML fragment in Javascript:

  \begin{lstlisting}[language=XML]
<ul>
  <li>een</li>
  <li>twee</li>
</ul>
  \end{lstlisting}

\end{frame}


\subsection{Existing Methods}

\begin{frame}[fragile]

  \frametitle{Generating String Literals}

  \begin{block}{Using string literals}
    \begin{lstlisting}[language=Java]
var menu =
   "<ul>"
 + "<li>een</li>"
 + "<li>twee</li>"
 + "</ul>";
    \end{lstlisting}
  \end{block}

\end{frame}


\begin{frame}[fragile]

  \frametitle{Manipulating Abstract Syntax Trees}

  \begin{block}{Using abstract syntax}
    \begin{lstlisting}[language=Java]
var menu, item;
menu = document.createElement("ul");

item = document.createElement("li");
item.appendChild(document.createTextNode("een"));
menu.appendChild(item);

item = document.createElement("li");
item.appendChild(document.createTextNode("twee"));
menu.appendChild(item);
    \end{lstlisting}
  \end{block}

\end{frame}


\begin{frame}

  \frametitle{Comparing String Literals and Astract Syntax}

  \begin{block}{Generating String Literals}
    \begin{description}
      \item{\color{darkgreen}$\Rightarrow$} Easy to read/write
      \item{\color{darkgreen}$\Rightarrow$} Easy to implement
      \item{\color{darkred}$\Rightarrow$} No syntactic checks
      \item{\color{darkred}$\Rightarrow$} Further manipulation will be hard
    \end{description}
    Conclusion: {\bf bad}
  \end{block}

  \begin{block}{Manipulating Abstract Syntax Trees}
    \begin{description}
      \item{\color{darkgreen}$\Rightarrow$} No syntactically incorrect code
      \item{\color{darkgreen}$\Rightarrow$} Further manipulation possible
      \item{\color{darkgreen}$\Rightarrow$} Matching on structure possible
      \item{\color{darkred}$\Rightarrow$} Knowledge of abstract syntax necessary
      \item{\color{darkred}$\Rightarrow$} Does not scale
    \end{description}
    Conclusion: {\bf `less bad'}
  \end{block}

\end{frame}


\subsection{Concrete Object Syntax}

\begin{frame}

  \frametitle{Concrete Syntax for Meta-Programming}

  \begin{block}{Solution}
    \begin{itemize}
      \item Use concrete syntax of the object language...
      \item ... but transform to abstract syntax manipulation.
    \end{itemize}
  \end{block}

  \begin{block}{Concrete object syntax}
    \begin{description}
      \item{\color{darkgreen}$\Rightarrow$} Easy to read/write
      \item{\color{darkgreen}$\Rightarrow$} No syntactically incorrect code
      \item{\color{darkgreen}$\Rightarrow$} Further manipulation possible
      \item{\color{darkgreen}$\Rightarrow$} Matching on structure possible
    \end{description}
    Conclusion: {\bf best of both worlds}
  \end{block}

\end{frame}


\begin{frame}[fragile]

  \frametitle{Example: Generating XML in Javascript}

  \begin{block}{Using concrete syntax}
    \begin{lstlisting}[language=Java]
var menu = <ul>
             <li>een</li>
             <li>twee</li>
           </ul>;
    \end{lstlisting}
  \end{block}

\end{frame}


\section{Implementation}

\frame{

  \frametitle{Concrete Syntax for Meta-Programming}

  \tableofcontents[currentsection]

}


\begin{frame}

  \frametitle{Implementation}

  \begin{block}{Syntax Definition}
    \begin{itemize}
      \item Embed object language in meta-language.
      \item Define one combined syntax.
    \end{itemize}
  \end{block}

  \begin{block}{Assimilation}
    \begin{itemize}
      \item Rewrite concrete object syntax to meta-language constructs.
      \item By program transformation.
    \end{itemize}
  \end{block}

\end{frame}


\subsection{Syntax Definition}

\begin{frame}

  \frametitle{Syntax Definition Formalism}

  \begin{block}{SDF2...}
    \begin{itemize}
      \item is declarative.
      \item is modular.
      \item combines lexical and context-free syntax.
    \end{itemize}
  \end{block}

  \begin{block}{Parsing}
    SGLR is a parser generator for SDF2 syntax definitions.
    \begin{itemize}
      \item Scannerless.
      \item Accepts full class of context-free grammars.
    \end{itemize}
  \end{block}

\end{frame}


\begin{frame}[fragile]

  \frametitle{Example: Embedding XML in Javascript}

  \begin{itemize}
    \item Syntax definition for Javascript (\texttt{Javascript})
    \item Syntax definition for XML (\texttt{Xml})
  \end{itemize}

  \begin{block}{Combined Syntax Definition}
    \begin{lstlisting}[language=SDF]
module JavascriptXml
imports Xml
imports Javascript

exports
  context-free syntax
    Element  ->  Expr {cons("ToExpr")}
    \end{lstlisting}
  \end{block}

\end{frame}


\subsection{Assimilation}

\begin{frame}

  \frametitle{Stratego}

  \begin{block}{Stratego language}
    \begin{itemize}
      \item Transformation language
      \item Rewrite object fragments to meta-language constructs
      \item Supports concrete object syntax itself
    \end{itemize}
  \end{block}

\end{frame}


\begin{frame}[fragile]

  \frametitle{Example: Assimilation of XML into Javascript}

  \begin{itemize}
    \item Look for fragments of concrete XML code
    \item Transform into AST manipulating Javascript code
  \end{itemize}

  \begin{lstlisting}
var t = <p>test</p>;
  \end{lstlisting}

$\Rightarrow$

  \begin{lstlisting}
var t;
{
  var _t = document.createElement('p');
  var _c = document.createTextNode('test');
  _t.appendChild(_c);
  t = _t
}
  \end{lstlisting}

\end{frame}


\section{Outro}

\frame{

  \frametitle{Concrete Syntax for Meta-Programming}

  \tableofcontents[currentsection]

}


\begin{frame}

  \frametitle{Conclusions}

  \begin{block}{Concrete object syntax}
    \begin{description}
      \item{\color{darkgreen}$\Rightarrow$} Much easier to write and read
      \item{\color{darkgreen}$\Rightarrow$} Usual strengths of AST manipulation
      \item{\color{darkred}$\Rightarrow$} Complicates programming environment
    \end{description}
  \end{block}

  \begin{block}{Future work}
    \begin{itemize}
      \item Integrate assimilation in meta-language compiler
      \item Apply method to library interfacing
      \item ...
    \end{itemize}
  \end{block}

\end{frame}


\begin{frame}

  \frametitle{Questions and Further Reading}

  Any questions?\\[4em]

  \begin{block}{Further Reading}
    \texttt{http://metaborg.org}
  \end{block}

\end{frame}


\end{document}
