\documentclass{beamer}

\usepackage[english]{babel}

\usepackage[latin1]{inputenc}

\usepackage[T1]{fontenc}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}

%\usepackage{tgtermes}
\usepackage{tgheros}
%\usepackage{cmbright}
\usepackage{qtxmath}
%\usepackage{kpfonts}
\renewcommand{\ttdefault}{txtt}

%\usepackage[amsmath,thmmarks]{ntheorem}

\usepackage{beamerthemesplit}

\setbeamertemplate{background canvas}[vertical shading][bottom=red!10,top=blue!10]
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{headline}{}
\usetheme{Warsaw}
\useinnertheme{rectangles}

\colorlet{darkred}{red!80!black}
\colorlet{darkblue}{blue!80!black}
\colorlet{darkgreen}{green!80!black}

\title{SVEP: A sequence variant effect predictor}

\author{Martijn Vermaat}
\institute{Human Genetics, LUMC}
\date{Bioinformaticians work discussion January 3, 2011}


\begin{document}


\frame{\titlepage}


\frame{

  \frametitle{SVEP: A sequence variant effect predictor}

  \tableofcontents

}


\section{Context}


\frame{

  \frametitle{SVEP: A sequence variant effect predictor}

  \tableofcontents[currentsection]

}


\frame{

  \frametitle{Effect prediction for variants}

  Identify or predict:
  \begin{itemize}
    \item Splice sites
    \item Exonic splicing enhancers
    \item Branch points
    \item Transcription factor binding sites
    \item Protein level changes
    \item $\ldots$
  \end{itemize}

  \vskip10pt

  \pause

  Many tools exist:
  \begin{itemize}
    \item Online, for local installation, or both
    \item Different models and algorithms
    \item Different input/output formats
    \item Different use cases (with overlap)
    \item $\ldots$
  \end{itemize}

}


\frame{

  \frametitle{Problems}

  Problems with using these tools:
  \begin{itemize}
    \item Which tools exist?
    \item Where are they?
    \item Which tools should I use?
    \item How can I compare tool results?
    \item $\ldots$
  \end{itemize}

  \vskip10pt

  \pause

  \begin{block}{Short version}
    It takes a lot of time.
  \end{block}

}


\section{SVEP}


\frame{

  \frametitle{SVEP: A sequence variant effect predictor}

  \tableofcontents[currentsection]

}


\frame{

  \frametitle{SVEP: A sequence variant effect predictor}

  \begin{block}{Basic idea}
    We want to automate:
    \begin{enumerate}
      \item Deciding which tools are interesting
      \item Running these tools
      \item Analysing their results
      \item Drawing conclusions
    \end{enumerate}
  \end{block}

}


\frame{

  \frametitle{1: Deciding which tools are interesting}

  \begin{block}{Might be a recursive process}
    \begin{enumerate}
      \item Choose tool(s) to run
      \item Run tool(s)
      \item Goto \textbf{1}
    \end{enumerate}
  \end{block}

  \vskip10pt

  \pause

  Based on
  \begin{itemize}
    \item heuristics from biologists
    \item statistics from results with a test set
    \item static tool properties (organism, \# base pairs, etc)
  \end{itemize}
  we can assign weights to all tools.\\[1em]

  \pause

  Basically: construct a decision tree.

}


\frame{

  \frametitle{2: Running these tools}

  Interface with:
  \begin{itemize}
    \item Webservices (preferably)
    \item Websites (HTML scraping results)
    \item Locally running tools
  \end{itemize}

  \vskip10pt

  \pause

  We have to deal with:
  \begin{itemize}
    \item Long (or very long) running tools
    \item Failing tools
    \item Tool limitations
  \end{itemize}

}


\frame{

  \frametitle{3: Analysing their results}

  \begin{itemize}
    \item Normalize
    \item Compare
    \item Present
    \item $\ldots$
  \end{itemize}

}


\frame{

  \frametitle{4: Drawing conclusions}

  (Needs input from biologists.)

}


\section{Implementation}


\frame{

  \frametitle{SVEP: A sequence variant effect predictor}

  \tableofcontents[currentsection]

}


\frame{

  \frametitle{Existing SVEP implementation}

  By Tessa Dorival (2007) and Miranda ter Laak (2009).

  \vskip10pt

  Written in Python with a frontend in PHP.

  \vskip10pt

  \begin{itemize}
    \item Interfaces $\approx 10$ tools by HTML scraping
    \item Runs all tools sequentially
    \item Does not use Mutalyzer
    \item Unfinished
  \end{itemize}

}


\frame{

  \frametitle{New SVEP implementation}

  Mostly a rewrite by me (December) in Python.

  \vskip10pt

  \begin{itemize}
    \item Interfaces $\approx 14$ tools by HTML scraping and SOAP
    \item Runs tools in parallel
    \item Stores processed tool results
    \item Unfinished
  \end{itemize}

  \vskip10pt

  \pause

  Run many jobs and store heterogeneous data:
  \begin{itemize}
    \item gevent: micro threads on top of libevent
    \item MongoDB: high-performance schema-less document store
  \end{itemize}

}


\end{document}
